---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fda.vi

<!-- badges: start -->
[![R-CMD-check](https://github.com/steviek16/fda.vi/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/steviek16/fda.vi/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**fda.vi** is an R package that implements a variational expectation-maximization (VEM) algorithm
for smoothing one or multiple functional observations via basis function
selection. The algorithm estimates all model parameters simultaneously and
automatically, while accounting for within-curve correlation. It provides a
fast, scalable alternative to Bayesian sampling-based MCMC methods for smoothing
functional curves with within-curve correlated errors via an Ornstein-Uhlenbeck
covariance function.

The details of the variational EM algorithm can be found in da Cruz, de Souza
and Sousa (2024). The package was developed as part of the Data Science 4999Z
Thesis at Western University.

## Installation
```r
# install.packages("devtools")
devtools::install_github("steviek16/fda.vi")
```

## Quick Start
```{r example-fit, message = FALSE}
library(fda.vi)

data(toy_curves)
set.seed(1234)

fit <- vem_fit(
  y      = toy_curves$y,
  Xt     = toy_curves$Xt,
  K      = 8,
  center = FALSE,
  scale  = FALSE
)
```

```{r example-summary}
summary(fit)
```

```{r example-plot}
plot(fit, curve_idx = 1)
```

```{r example-coef}
coef(fit)
```

```{r example-predict}
Xt_new <- seq(0, 1, length.out = 200)
predict(fit, newdata = Xt_new)
```

## Key Features

- **Automatic Bayesian Basis Selection** via a sparsity-inducing prior on basis
  coefficients
- **Uncertainty Quantification** via 95% credible bands, constructed by
  posteriors of the basis coefficients and inclusion indicators
- **Correlated (Within-Curve) Error Structure** modelled explicitly using an
  Ornstein-Uhlenbeck covariance function, with the decay parameter estimated
  automatically via the M-step
- **Automatic Selection of Number of Basis Functions K** via generalized cross-validation (GCV) over
  a user-supplied grid of candidate basis sizes
- **Multiple Basis Types** supported: cubic B-splines and Fourier bases
- **S3 User Methods**: `plot()`, `predict()`, `coef()`, and `summary()`
  methods for fitted objects
- **Fast** The variational EM algorithm converges in tens of iterations

## References

da Cruz, A. C., de Souza, C. P. E., & Sousa, P. H. T. O. (2024).
Fast Bayesian basis selection for functional data representation with
correlated errors. *arXiv:2405.20758*.
<https://arxiv.org/abs/2405.20758>

## Citation

To cite the method implemented in this package, please use:

da Cruz, A. C., de Souza, C. P. E., & Sousa, P. H. T. O. (2024).
Fast Bayesian basis selection for functional data representation with
correlated errors. *arXiv:2405.20758*.
<https://arxiv.org/abs/2405.20758>

```bibtex
@misc{dacruz2024vem,
  title     = {Fast {Bayesian} basis selection for functional data
               representation with correlated errors},
  author    = {da Cruz, Ana Carolina and de Souza, Camila P. E. and
               Sousa, Pedro H. T. O.},
  year      = {2024},
  note      = {arXiv:2405.20758},
  url       = {https://arxiv.org/abs/2405.20758}
}
```



